From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-09 Pi+/paper/discussion.tex
more on sec 8

The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
presentation of reversible programming based on algebraic types matches---exactly---the categorified group-theoretic
semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible
languages. Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence is
established in this work.

@@

@@

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua_\beta} : (e : A \eqv B) \to
    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua_\eta} : (p : A \id_{\UU} B) \to p ==
    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
$\term{ua}-\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.

By giving a computable presentation for a univalent subuniverse, we are able to describe the path space of it
syntactically, by giving a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
The $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator), which is easily done by induction.
The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which, in general, is of course
undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean circuits, it is decidable, as
we have shown, but still far from trivial, which matches the need to assert the existence of $\term{ua}$ without giving
a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact that program synthesis is
sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$ 2-combinators). Thus, we
this suggests a new computational interpretation of the univalence principle, which provides an intuition on why certain
constructions are hard (or impossible in the general case).

We could present a dependent type theory for the topos $\SetFin$, with an identity type for terms (generated by
$\refl$), and one for types (generated by $\PiLang$ combinators). We can't talk about universes in $\SetFin$ since it
doesn't have one, but we could show externally that it satisfies univalence.

%%%%%%%%%%%%%%
\paragraph{Univalent Fibrations} Univalent Fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018},
to build a model of Voevodsky's \emph{univalence} principle in simplicial sets.
\citet{christensenCharacterizationUnivalentFibrations2015} studied characterisations of univalent fibrations using the
$\BAut$ construction. Univalent typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent
subuniverses.

Coherence problems in type theory, coherence via Well-Foundedness.

Formalised proofs of Mac Lane's coherence theorem.

Applications of FSMG and history of the coherence theorems.

Other proofs of coherence theorems, Joyal-Street.

Using our presentation of $S_{n}$, we can construct the Eilenberg-Maclane space (using a HIT) $K(S_{n},1)$. Then, it
should be true that $\UFin \eqv \sqcup_{n:\Nat} K(S_{n},1)$. This is future work.

Other applications of symmetric groups.

Actions of symmetric groups, permutation groupoids.

>>>>

The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
presentation of reversible programming based on algebraic types matches---exactly---the categorified group-theoretic
semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible
languages. Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence
suggested by~\citet{Carette2016} and~\citet{threemodels} and established in this work.

@@

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua_\beta} : (e : A \eqv B) \to
    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua_\eta} : (p : A \id_{\UU} B) \to p ==
    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
$\term{ua}_\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.

By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
via a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
which is easily done by induction.  The $\term{ua}$ map corresponds to synthesising a program from an equivalence
(which, in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible
boolean circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the
existence of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses
the fact that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory
($\PiLang$ 2-combinators). Thus, our results suggests a new computational interpretation of the univalence principle, which
provides an intuition on why certain constructions are hard (or impossible in the general case).

% We could present a dependent type theory for the topos $\SetFin$, with an identity type for terms (generated by
% $\refl$), and one for types (generated by $\PiLang$ combinators). We can't talk about universes in $\SetFin$ since it
% doesn't have one, but we could show externally that it satisfies univalence.

Univalent Fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018},
to build a model of Voevodsky's \emph{univalence} principle in simplicial sets.
\citet{christensenCharacterizationUnivalentFibrations2015} studied characterisations of univalent fibrations using the
$\BAut$ construction. Univalent typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent
subuniverses.

%%%%%%%%%%%%%%
% \paragraph{Univalent Fibrations}

% Coherence problems in type theory, coherence via Well-Foundedness.

% Formalised proofs of Mac Lane's coherence theorem.

% Applications of FSMG and history of the coherence theorems.

% Other proofs of coherence theorems, Joyal-Street.

% Pi has other extensions (fractional/negative/recursive types).
% What are the free X monoidal structures they're describing?

% Using our presentation of $S_{n}$, we can construct the Eilenberg-Maclane space (using a HIT) $K(S_{n},1)$. Then, it
% should be true that $\UFin \eqv \sqcup_{n:\Nat} K(S_{n},1)$. This is future work.

% Other applications of symmetric groups.

% Actions of symmetric groups, permutation groupoids.

%%%%%%%%%%%%%%
\paragraph{Reversible Programming Languages}

@@
